The Unicode Cookbook \linebreak for Linguists

### 2016-08-18 book/chapters/implementation.tex
typos

\noindent The version available through CRAN is the latest stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development version, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

@@

\noindent Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called `vignette' with
explanations and examples.

@@

\noindent Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

@@

\noindent For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

@@

\noindent After this softlink it should be possible to access the \texttt{tokenize}
function from the shell. Try \texttt{tokenize --help} to test the functionality.

<<size='scriptsize', engine='bash'>>=
tokenize --help
@

@@

The first example of how to use these functions considers finding errors in the
encoding of texts. In the following example, it looks as if we have two
identical strings, \texttt{AABB}. However, this is just an surface-impression
delivered by the current font, which renders Latin and Cyrillic capitals
identically. We can identify this problem when we produce an orthography profile
from the strings. Using here the R implementation of orthography profiles, we
first assign the two strings to a variable \texttt{test}, and then produce an
orthography profile with the function \texttt{write.profile}. As it turns out,
some of the letters are Cyrillic.

@@

\noindent The function of error-message reporting can also nicely be illustrated
with this example. Suppose we made an orthography profile with just the two
Latin letters <A> and <B> as possible graphemes, then this profile would not be
sufficient to tokenize the strings. There are graphemes in the data that are not
in the profile, so the tokenization produces an error, which can be used to fix
the encoding (or the profile). In the example below, we can see that the
Cyrillic encoding is found in the second string of the \texttt{test} input.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
turns out, ome of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

\noindent By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

@@

\noindent Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

@@

\noindent It is important to realize that for Unicode grapheme definitions, NFC
and NFD normalization are equivalent. This can be shown by normalizing the
example in either NFC or NFD, but using default separation in
\texttt{write.profile}. To be precise, default separation means setting
\texttt{sep~=~NULL}, but that has not be added explicitly.

@@

\noindent Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see \ref{available-implementations} for notes
on how to install the bash executable). Exactly this example is also included in
the help file of the executable.

<<size='scriptsize', engine='bash'>>=
## show help
writeprofile --help
@

@@

\noindent It is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simply situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

\noindent This will lead to the following four files being written. Crucially, a
new profile is produced with the re-ordered orthography profile. To reproduce
the tokenization, this re-ordered profile can be used with the option
\texttt{ordering~=~NULL}.

@@

Besides the columns \texttt{Left}, \texttt{Grapheme}, and \texttt{Right} as
discussed in the previous sections, there are also columns \texttt{Class} and
\texttt{Replacement}. The column \texttt{Class} can be used to specify classes
of graphemes that can then be used in the contextual specification. The column
\texttt{Replacement} is just a copy of the column \textt{Grapheme}, providing a
skeleton to specify transliteration. The name of the columne
\texttt{Replacement} is not fixed -- there can actually be multiple columns with 
different kinds of transliterations in a single profile.

@@

In all examples above we have used the default formatting for grapheme
separation using space as a separator, which is obtained by the default setting
\texttt{sep = " "}. It is possible to specify any other separator here,
including the empty string, i.e. \texttt{sep = ""}. This will not show the
graphemic tokenization anymore (although it has of course been used in the
background).

@@

Given a set of graphemes, there are at least two different methods to tokenize
strings. The first is called \texttt{method = "global"} here: this approach
takes the first grapheme in the profile, then matches this grapheme globally at
all places in the string, and then turns to the next string in the profile. To
us, this seems to be the most intuitive way for linguists to deal with graphemic
parsing. The other approach is called \texttt{method = "linear"} here: this
approach walks through the string from left to right. At the first character it
looks through all graphemes whether there is any match, and then walks further
to the end of the match and starts again. This approach is more alike to
finite-state rewrite rules (though note that it still works differently from
such rewrite rules, see Section~\ref{rule-ordering}).

In some special cases these two methods can lead to different results, but these
special situations are very unlikely to happen in natural language. The example
below shows that a string \texttt{'abc'} can be parsed differently in case of a
very special profile with a very special ordering of the graphemes.

<<tidy=FALSE>>=
# different parsing methods can lead to different results
# the global method first catches 'bc'
tokenize( "abc"
         , profile = c("bc","ab","a","c")
         , order = NULL
         , method = "global"
         )$strings
# the linear method catches the first grapheme, which is 'ab'
tokenize( "abc"
         , profile = c("bc","ab","a","c")
         , order = NULL
         , method = "linear"
         )$strings
@

The current R-implementation has a limitation when regular expressions are used. 
The problem is that overlapping matches are not captured. Everything works as 
expected without regular expressions, but there might be warnings/errors in case of 
\texttt{regex = TRUE}. However, just as in the previous issue, this should very 
rarely happen in natural language.

@@

<<tidy=FALSE>>=
# Everything perfekt without regular expressions
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = FALSE
        )$strings
# Matching with regular expressions does not catch overlap
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = TRUE
        )$strings
@

>>>>

The version available through CRAN is the latest stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development version, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

@@

Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called `vignette' with
explanations and examples.

@@

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

@@

For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

@@

After inserting this softlink it should be possible to access the
\texttt{tokenize} function from the shell. Try \texttt{tokenize --help} to test
the functionality.

<<size='scriptsize', engine='bash', tidy=FALSE>>=
tokenize --help
@

@@

The first example of how to use these functions concerns finding errors in the
encoding of texts. In the following example, it looks as if we have two
identical strings, \texttt{AABB}. However, this is just an surface-impression
delivered by the current font, which renders Latin and Cyrillic capitals
identically. We can identify this problem when we produce an orthography profile
from the strings. Using here the R implementation of orthography profiles, we
first assign the two strings to a variable \texttt{test}, and then produce an
orthography profile with the function \texttt{write.profile}. As it turns out,
some of the letters are Cyrillic.

@@

The function of error-message reporting can also nicely be illustrated
with this example. Suppose we made an orthography profile with just the two
Latin letters <A> and <B> as possible graphemes, then this profile would not be
sufficient to tokenize the strings. There are graphemes in the data that are not
in the profile, so the tokenization produces an error, which can be used to fix
the encoding (or the profile). In the example below, we can see that the
Cyrillic encoding is found in the second string of the \texttt{test} input.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

@@

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

@@

It is important to realize that for Unicode grapheme definitions, NFC
and NFD normalization are equivalent. This can be shown by normalizing the
example in either NFC or NFD, but using default separation in
\texttt{write.profile}. To be precise, default separation means setting
\texttt{sep~=~NULL}, but that has not be added explicitly.

@@

Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see \ref{available-implementations} for notes
on how to install the bash executable). Exactly this example is also included in
the help file of the executable.

<<size='scriptsize', engine='bash'>>=
writeprofile --help
@

@@

It is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simple situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

This will lead to the following four files being written. Crucially, a
new profile is produced with the re-ordered orthography profile. To reproduce
the tokenization, this re-ordered profile can be used with the option
\texttt{ordering~=~NULL}.

@@

Besides the columns \texttt{Left}, \texttt{Grapheme}, and \texttt{Right} as
discussed in the previous sections, there are also columns \texttt{Class} and
\texttt{Replacement}. The column \texttt{Class} can be used to specify classes
of graphemes that can then be used in the contextual specification. The column
\texttt{Replacement} is just a copy of the column \textt{Grapheme}, providing a
skeleton to specify transliteration. The name of the column
\texttt{Replacement} is not fixed -- there can actually be multiple columns with 
different kinds of transliterations in a single profile.

@@

In all examples above we have used the default formatting for grapheme
separation using space as a separator, which is obtained by the default setting
\texttt{sep~=~"~"}. It is possible to specify any other separator here,
including the empty string, i.e. \texttt{sep = ""}. This will not show the
graphemic tokenization anymore (although it has of course been used in the
background).

@@

Given a set of graphemes, there are at least two different methods to tokenize
strings. The first is called \texttt{method = "global"}. This approach
takes the first grapheme in the profile, then matches this grapheme globally at
all places in the string, and then turns to the next string in the profile. To
us, this seems to be the most intuitive way for linguists to deal with graphemic
parsing. The other approach is called \texttt{method = "linear"}. This
approach walks through the string from left to right. At the first character it
looks through all graphemes whether there is any match, and then walks further
to the end of the match and starts again. This approach is more alike to
finite-state rewrite rules (though note that it still works differently from
such rewrite rules, see Section~\ref{rule-ordering}). The global method is used 
by default.

In some special cases these two tokenization methods can lead to different
results, but these special situations are very unlikely to happen in natural
language. The example below shows that a string \texttt{'abc'} can be parsed
differently in case of a very special profile with a very special ordering of
the graphemes.

<<tidy=FALSE>>=
# different parsing methods can lead to different results
# the global method first catches 'bc'
tokenize( "abc"
         , profile = c("bc","ab","a","c")
         , order = NULL
         , method = "global"
         )$strings

# the linear method catches the first grapheme, which is 'ab'
tokenize( "abc"
         , profile = c("bc","ab","a","c")
         , order = NULL
         , method = "linear"
         )$strings
@

Further, the current R-implementation has a limitation when regular expressions
are used. The problem is that overlapping matches are not captured when using
regular expressions.\footnote{This restriction is an effect of the underlyingly
used ICU implementation of the Unicode standard as implemented in R through the
package \texttt{stringi}.} Everything works as expected without regular
expressions, but there might be warnings/errors in case of \texttt{regex =
TRUE}. However, just as in the previous issue, this problem should only very
rarely (when at all) happen in natural language data.

@@

<<tidy=FALSE>>=
# Everything perfekt without regular expressions
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = FALSE
        )$strings

# Matching with regular expressions does not catch overlap
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = TRUE
        )$strings
@
